---
title: "STA457 Final ProjectMultivariate Analysis"
output: pdf_document
date: "2025-04-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load the necessary libraries
library(tidyverse)
library(ggplot2)
library(forecast)
library(zoo)
library(dplyr)
library(lubridate)
library(timeSeries)
library(fGarch)
library(xts)
library(rugarch)
library(vars)
library(ggplot2)
library(tseries)
library(astsa)
```



# We merge the datasets, show joint stationarity
```{r}
# Load pre-processed monthly data, both created from STA457_Preprocessing_Submission.Rmd
cocoa_monthly <- read.csv("cocoa_monthly.csv")
ghana_monthly <- read.csv("ghana_monthly.csv")

cocoa_monthly$YearMonth <- as.Date(cocoa_monthly$YearMonth, format = "%Y-%m-%d")
ghana_monthly$YearMonth <- as.Date(ghana_monthly$YearMonth, format = "%Y-%m-%d")

# Log and difference the dataset
log_prices <- cocoa_monthly %>%
  group_by(YearMonth) %>%
  summarise(log_price = log(avg_price), na.rm = TRUE)

log_diff_cocoa <- log_prices %>%
  mutate(diff_price = c(NA, diff(log_price))) %>%
  group_by(YearMonth)

log_diff_cocoa$YearMonth <- as.Date(log_diff_cocoa$YearMonth, format = "%Y-%m-%d")
ghana_monthly$YearMonth <- as.Date(ghana_monthly$YearMonth, format = "%Y-%m-%d")

# Create separate monthly total precipitation and price dataset
monthly_prec <- ghana_monthly %>%
  mutate(YearMonth = floor_date(YearMonth, "month")) %>%
  group_by(YearMonth) %>%
  summarise(TotalPRCP)

# Create separate monthly average temperature and price dataset
monthly_temp <- ghana_monthly %>%
  mutate(YearMonth = floor_date(YearMonth, "month")) %>%
  group_by(YearMonth) %>%
  summarise(MeanTAVG)


# We merge the datasets
multivariate_temp <- merge(log_diff_cocoa, monthly_prec, by = "YearMonth", all = FALSE)

# Full, multivariate data
multivariate_cocoa <- merge(multivariate_temp, monthly_temp, by = "YearMonth", all = FALSE)
print(multivariate_cocoa)

# Remove the NA
multivariate_cocoa <- multivariate_cocoa %>%
  dplyr::filter(!is.na(diff_price))

adf.test(multivariate_cocoa$diff_price)
adf.test(multivariate_cocoa$TotalPRCP)
adf.test(multivariate_cocoa$MeanTAVG)

```
# Plot and heteroscedasticity measures for EDA
```{r}
# Plot average temperature and total precipitation values
plot(ghana_monthly$MeanTAVG,
     type="l")
plot(ghana_monthly$TotalPRCP,
     type="l")

n_ghana = nrow(ghana_monthly)

# Temperature Heteroscedasticity analysis
# Split into half
first_half_temp <- ghana_monthly$MeanTAVG[1:(n_ghana/2)]
second_half_temp <- ghana_monthly$MeanTAVG[(n_ghana/2 + 1):n_ghana]

# Compute sample variances for each half
var_first_temp <- var(first_half_temp)
var_second_temp <- var(second_half_temp)

# Print the sample variances for each half
cat("Sample variance for the first half of the mean temperature:", var_first_temp, "\n")
cat("Sample variance for the second half of the mean temperature:", var_second_temp, "\n")

# Split into last 2 years and the rest
split_temp <- ghana_monthly$MeanTAVG[1:(n_ghana-26)]
last_two_temp <- ghana_monthly$MeanTAVG[((n_ghana-26) + 1):n_ghana]

# Compute into last 2 years and the rest
var_split_temp <- var(split_temp)
var_last_two_temp <- var(last_two_temp)

# Print the sample variances
cat("Sample variance until the last two years of mean temperature:", var_split_temp, "\n")
cat("Sample variance since 2023 of mean temperature:", var_last_two_temp, "\n")



# Precipitation Heteroscedasticity analysis
# Split into half
first_half_prec <- ghana_monthly$TotalPRCP[1:(n_ghana/2)]
second_half_prec <- ghana_monthly$TotalPRCP[(n_ghana/2 + 1):n_ghana]

# Compute sample variances for each half
var_first_prec <- var(first_half_prec)
var_second_prec <- var(second_half_prec)

# Print the sample variances for each half
cat("Sample variance for the first half of the total precipitation:", var_first_prec, "\n")
cat("Sample variance for the second half of the total precipitation:", var_second_prec, "\n")

# Split into last 2 years and the rest
split_prec <- ghana_monthly$TotalPRCP[1:(n_ghana-26)]
last_two_prec <- ghana_monthly$TotalPRCP[((n_ghana-26) + 1):n_ghana]

# Compute into last 2 years and the rest
var_split_prec <- var(split_prec)
var_last_two_prec <- var(last_two_prec)

# Print the sample variances
cat("Sample variance until the last two years of total precipitation:", var_split_prec, "\n")
cat("Sample variance since 2023 of total precipitation:", var_last_two_prec, "\n")

```

## Mean Tempersture does not show significant changing variance over time

# Cross-correlation plots for dependance orders
```{r}
# cross correlation functions
ccf_temp_avg <- ccf(multivariate_cocoa$diff_price, multivariate_cocoa$MeanTAVG, lag = 60)
ccf_prcp <- ccf(multivariate_cocoa$diff_price, multivariate_cocoa$TotalPRCP, lag = 60)

# Extract the correlation values (acf) and the corresponding lags 
corr_values_temp <- as.numeric(ccf_temp_avg$acf)
lags_tmpavg <- ccf_temp_avg$lag

corr_values_prcp <- as.numeric(ccf_prcp$acf)
lags_prcp <- ccf_prcp$lag

# Find the index of the largest correlation in absolute value
id_tmpavg <- which.max(abs(corr_values_temp))
id_prcp <- which.max(abs(corr_values_prcp))

# Retrieve the correlation and the corresponding lag
largest_corr_temp <- corr_values_temp[id_tmpavg]
largest_lag_temp <- lags_tmpavg[id_tmpavg]

largest_corr_prcp <- corr_values_prcp[id_prcp]
largest_lag_prcp <- lags_prcp[id_prcp]

cat("The largest absolute correlation for the mean temperature is", largest_corr_temp, "at lag", largest_lag_temp, "\n")

cat("The largest absolute correlation for the total precipitation is", largest_corr_prcp, "at lag", largest_lag_prcp, "\n")
```
# Create new datasets with dependance orders in tact.
```{r}
# Create a new column with the time 6 months prior for each row 
log_diff_cocoa$lag_YearMonth_temp <- log_diff_cocoa$YearMonth %m-% months(6)

# Merge datasets match each cocoa price with average temperature from 6 months ago
df_temp <- merge(
  log_diff_cocoa[, c("YearMonth", "log_price", "lag_YearMonth_temp")],
  ghana_monthly[, c("YearMonth", "MeanTAVG")],
  by.x = "lag_YearMonth_temp",
  by.y = "YearMonth",
  all.x = TRUE
)

# Create a new column with the time 44 months prior for each row 
log_diff_cocoa$lag_YearMonth_prec <- log_diff_cocoa$YearMonth %m-% months(44)

# Merge datasets match each cocoa price with the total precipitation from 44 months ago
df_prec <- merge(
  log_diff_cocoa[, c("YearMonth", "log_price", "lag_YearMonth_prec")],
  ghana_monthly[, c("YearMonth", "TotalPRCP")],
  by.x = "lag_YearMonth_prec",
  by.y = "YearMonth",
  all.x = TRUE
)

# Remove NA values
df_temp <- df_temp %>%
  dplyr::filter(!is.na(MeanTAVG))

df_prec <- df_prec %>%
  dplyr::filter(!is.na(TotalPRCP))

# Preview the resulting datasets
print(df_temp)
print(df_prec)

```


```{r}
# Ljung-Box test function with standardized residuals
plot_ljung_box <- function(model, model_name = deparse(substitute(model)), max_lag = 20) {
  p_values <- sapply(1:max_lag, function(lag) Box.test(resid(model) / sqrt(model$sigma2),
  lag = lag, type = "Ljung-Box")$p.value)
  
  df <- data.frame(Lag = 1:max_lag, P_Value = p_values)
  
  ggplot(df, aes(x = Lag, y = P_Value)) +
    geom_point(shape = 1, size = 3) +  # Open circles
    geom_hline(yintercept = 0.05, linetype = "dashed", color = "blue") +  # Significance threshold
    ylim(0, 1) +
    labs(title = paste("p values for Ljung-Box statistic for", model_name),
         x = "LAG (H)",
         y = "p value") +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(size = 12)
    )
}
```

# ARIMAX model implementation
```{r}
# ARIMAX with 6-step behind average temperature 
## ARIMA(1,1,0)
arimax1_temp <- arima(df_temp$log_price, c(1,1,0), xreg = df_temp$MeanTAVG)
## ARIMA(0,1,1)
arimax2_temp <- arima(df_temp$log_price, c(0,1,1), xreg = df_temp$MeanTAVG)
## ARIMA(6,1,7)
arimax3_temp <- arima(df_temp$log_price, c(6,1,7), xreg = df_temp$MeanTAVG)

# ARIMAX with 44-step behind total precipitation
## ARIMA(1,1,0)
arimax1_prec <- arima(df_prec$log_price, c(1,1,0), xreg = df_prec$TotalPRCP)
## ARIMA(0,1,1)
arimax2_prec <- arima(df_prec$log_price, c(0,1,1), xreg = df_prec$TotalPRCP)
## ARIMA(6,1,7)
arimax3_prec <- arima(df_prec$log_price, c(6,1,7), xreg = df_prec$TotalPRCP)

# Residual Analysis of temperature ARIMAX models
## Ljung-Box Plots
plot_ljung_box(arimax1_temp)
plot_ljung_box(arimax2_temp)
plot_ljung_box(arimax3_temp)
## Retrieve Standardized Residuals
stand_arimax_AR = resid(arimax1_temp) / sqrt(arimax1_temp$sigma2)
stand_arimax_MA = resid(arimax2_temp) / sqrt(arimax2_temp$sigma2)
stand_arimax_ARMA = resid(arimax3_temp) / sqrt(arimax3_temp$sigma2)
## QQ plot for AR component ARIMAX
qqnorm(stand_arimax_AR)
qqline(stand_arimax_AR, col = "red")
## QQ plot for MA component ARIMAX
qqnorm(stand_arimax_MA)
qqline(stand_arimax_MA, col = "blue")
## QQ plot for ARMA component ARIMAX
qqnorm(stand_arimax_ARMA)
qqline(stand_arimax_ARMA, col = "green")
## Summary
summary(arimax1_temp) # aic = -972.52
summary(arimax2_temp) # aic = -974.00
summary(arimax3_temp) # aic = -969.20

# Residual Analysis of Precipitation ARIMAX models
## Ljung-Box Plots
plot_ljung_box(arimax1_prec)
plot_ljung_box(arimax2_prec)
plot_ljung_box(arimax3_prec)
## Retrieve Standardized Residuals
stand_arimax_AR_prec = resid(arimax1_prec) / sqrt(arimax1_prec$sigma2)
stand_arimax_MA_prec = resid(arimax2_prec) / sqrt(arimax2_prec$sigma2)
stand_arimax_ARMA_prec = resid(arimax3_prec) / sqrt(arimax3_prec$sigma2)

## QQ plot for AR component ARIMAX
qqnorm(stand_arimax_AR_prec)
qqline(stand_arimax_AR_prec, col = "red")
## QQ plot for MA component ARIMAX
qqnorm(stand_arimax_MA_prec)
qqline(stand_arimax_MA_prec, col = "blue")
## QQ plot for ARMA component ARIMAX
qqnorm(stand_arimax_ARMA_prec)
qqline(stand_arimax_ARMA_prec, col = "green")

## Summary
summary(arimax1_prec) # aic = -970.79
summary(arimax2_prec) # aic = -970.02
summary(arimax3_prec) # aic = -979.55

```
## Residual assumptions look satisfied for all models.

# Train/test plots and errors.
## Using ARIMAX (6,1,7) + temperature for train/test set.
```{r}
# Load necessary libraries so they do not get mixed up 
library(forecast)
library(stats) 
library(ggplot2)
library(lubridate)

# Ensure the data is sorted chronologically by YearMonth
df_temp <- df_temp[order(df_temp$YearMonth), ]

# Determine the start year and month for the time series
start_year <- year(df_temp$YearMonth[1])
start_month <- month(df_temp$YearMonth[1])

# Create time series objects for log_price and the external regressor
# Frequency is 12 for monthly data
log_price_ts <- ts(df_temp$log_price, start = c(start_year, start_month), frequency = 12)
mean_tavg_ts <- ts(df_temp$MeanTAVG, start = c(start_year, start_month), frequency = 12)

# Determine the total length and the forecast horizon (last 6 months)
n <- length(log_price_ts)
h6 <- 6 # Forecast horizon

# Create training sets using window()
train_log_price <- window(log_price_ts, end = time(log_price_ts)[n - h6])
train_xreg_ts   <- window(mean_tavg_ts, end = time(mean_tavg_ts)[n - h6])

# Create test sets using window()
test_log_price <- window(log_price_ts, start = time(log_price_ts)[n - h6 + 1])
test_xreg_ts   <- window(mean_tavg_ts, start = time(mean_tavg_ts)[n - h6 + 1])
test_dates     <- time(test_log_price) # Get numeric time index for plotting/dataframe

# Convert train_xreg_ts to matrix 
train_xreg_matrix <- as.matrix(train_xreg_ts)
colnames(train_xreg_matrix) <- c("MeanTAVG") 

# Fit the ARIMAX(6,1,7) model 
model_arimax_6m <- tryCatch({
  # Using forecast::Arima
  Arima(train_log_price, order = c(6, 1, 7), xreg = train_xreg_matrix)
}, error = function(e) {
  cat("Error fitting Arima model:", conditionMessage(e), "\n")
  return(NULL)
})

# Forecast on the Test Set ---
if (!is.null(model_arimax_6m)) {

  # Convert test_xreg_ts to matrix 
  test_xreg_matrix <- as.matrix(test_xreg_ts)
  colnames(test_xreg_matrix) <- c("MeanTAVG")
  fc6_arimax <- forecast(model_arimax_6m, h = h6, xreg = test_xreg_matrix)

  # Compute accuracy metrics using forecast::accuracy() on the log scale forecast object
  accuracy6_log <- accuracy(fc6_arimax, test_log_price)
  cat("ARIMAX(6,1,7) 6-Month Forecast Accuracy (Log Scale):\n")
  print(accuracy6_log)

    # Convert forecasts (mean) and actual test values back to the original price scale
  fc6_orig <- exp(fc6_arimax$mean)
  actual6_orig <- exp(test_log_price)

  # Compute error metrics manually on the original scale
  mae_orig <- mean(abs(fc6_orig - actual6_orig))
  rmse_orig <- sqrt(mean((fc6_orig - actual6_orig)^2))
  # Add a small epsilon to actual6_orig denominator to avoid division by zero if necessary
  mape_orig <- mean(abs((fc6_orig - actual6_orig) / (actual6_orig + 1e-10))) * 100

  # Print the results for the original scale
  cat("\nARIMAX(6,1,7) 6-Month Forecast Accuracy on Original Scale:\n")
  cat("MAE:", mae_orig, "\n")
  cat("RMSE:", rmse_orig, "\n")
  cat("MAPE:", mape_orig, "%\n")

# Create a data frame for plogginb
  df6_arimax <- data.frame(
    Month = as.numeric(test_dates), # Use the numeric time index
    Actual = as.numeric(actual6_orig),
    Forecast = as.numeric(fc6_orig)
  )

  # Create the plot 
  p6_arimax <- ggplot(df6_arimax, aes(x = Month)) +
    geom_line(aes(y = Actual, color = "Actual"), linewidth = 1) +
    geom_line(aes(y = Forecast, color = "Forecast"), linewidth = 1, linetype = "dashed") +
    scale_color_manual(name = "Legend", values = c("Actual" = "blue", "Forecast" = "red")) + # Manual colors
    labs(title = "ARIMAX(6,1,7) with Temperature 6-Month Forecast vs Actual",
         y = "Price (Original Scale)",
         x = "Time Index") + 
    theme_minimal() +
    theme(legend.position = "top")

  # Print the plot
  print(p6_arimax)

} else {
  cat("Skipping prediction, evaluation, and plotting due to model fitting error.\n")
}
```

## Using ARIMAX (1,1,0) + temperature for train/test set.
```{r}
# Load necessary libraries so they do not get mixed up 
library(forecast)
library(stats) 
library(ggplot2)
library(lubridate)

# Ensure the data is sorted chronologically by YearMonth
df_temp <- df_temp[order(df_temp$YearMonth), ]

# Determine the start year and month for the time series
start_year <- year(df_temp$YearMonth[1])
start_month <- month(df_temp$YearMonth[1])

# Create time series objects for log_price and the external regressor
# Frequency is 12 for monthly data
log_price_ts <- ts(df_temp$log_price, start = c(start_year, start_month), frequency = 12)
mean_tavg_ts <- ts(df_temp$MeanTAVG, start = c(start_year, start_month), frequency = 12)

# Determine the total length and the forecast horizon (last 6 months)
n <- length(log_price_ts)
h6 <- 6 # Forecast horizon

# Create training sets using window()
train_log_price <- window(log_price_ts, end = time(log_price_ts)[n - h6])
train_xreg_ts   <- window(mean_tavg_ts, end = time(mean_tavg_ts)[n - h6])

# Create test sets using window()
test_log_price <- window(log_price_ts, start = time(log_price_ts)[n - h6 + 1])
test_xreg_ts   <- window(mean_tavg_ts, start = time(mean_tavg_ts)[n - h6 + 1])
test_dates     <- time(test_log_price) # Get numeric time index for plotting/dataframe

# Convert train_xreg_ts to matrix 
train_xreg_matrix <- as.matrix(train_xreg_ts)
colnames(train_xreg_matrix) <- c("MeanTAVG") 

# Fit the ARIMAX(1,1,0) model 
model_arimax_6m <- tryCatch({
  # Using forecast::Arima
  Arima(train_log_price, order = c(1, 1, 0), xreg = train_xreg_matrix)
}, error = function(e) {
  cat("Error fitting Arima model:", conditionMessage(e), "\n")
  return(NULL)
})

# Forecast on the Test Set ---
if (!is.null(model_arimax_6m)) {

  # Convert test_xreg_ts to matrix 
  test_xreg_matrix <- as.matrix(test_xreg_ts)
  colnames(test_xreg_matrix) <- c("MeanTAVG")
  fc6_arimax <- forecast(model_arimax_6m, h = h6, xreg = test_xreg_matrix)

  # Compute accuracy metrics using forecast::accuracy() on the log scale forecast object
  accuracy6_log <- accuracy(fc6_arimax, test_log_price)
  cat("ARIMAX(1,1,0) 6-Month Forecast Accuracy (Log Scale):\n")
  print(accuracy6_log)

    # Convert forecasts (mean) and actual test values back to the original price scale
  fc6_orig <- exp(fc6_arimax$mean)
  actual6_orig <- exp(test_log_price)

  # Compute error metrics manually on the original scale
  mae_orig <- mean(abs(fc6_orig - actual6_orig))
  rmse_orig <- sqrt(mean((fc6_orig - actual6_orig)^2))
  # Add a small epsilon to actual6_orig denominator to avoid division by zero if necessary
  mape_orig <- mean(abs((fc6_orig - actual6_orig) / (actual6_orig + 1e-10))) * 100

  # Print the results for the original scale
  cat("\nARIMAX(1,1,0) 6-Month Forecast Accuracy on Original Scale:\n")
  cat("MAE:", mae_orig, "\n")
  cat("RMSE:", rmse_orig, "\n")
  cat("MAPE:", mape_orig, "%\n")

# Create a data frame for plogginb
  df6_arimax <- data.frame(
    Month = as.numeric(test_dates), # Use the numeric time index
    Actual = as.numeric(actual6_orig),
    Forecast = as.numeric(fc6_orig)
  )

  # Create the plot 
  p6_arimax <- ggplot(df6_arimax, aes(x = Month)) +
    geom_line(aes(y = Actual, color = "Actual"), linewidth = 1) +
    geom_line(aes(y = Forecast, color = "Forecast"), linewidth = 1, linetype = "dashed") +
    scale_color_manual(name = "Legend", values = c("Actual" = "blue", "Forecast" = "red")) + # Manual colors
    labs(title = "ARIMAX(1,1,0) 6-Month Forecast vs Actual (Original Scale)",
         y = "Price (Original Scale)",
         x = "Time Index") + 
    theme_minimal() +
    theme(legend.position = "top")

  # Print the plot
  print(p6_arimax)

} else {
  cat("Skipping prediction, evaluation, and plotting due to model fitting error.\n")
}
```

## Using ARIMAX (0,1,1) + temperature for train/test set.
```{r}
# Load necessary libraries so they do not get mixed up 
library(forecast)
library(stats) 
library(ggplot2)
library(lubridate)

# Ensure the data is sorted chronologically by YearMonth
df_temp <- df_temp[order(df_temp$YearMonth), ]

# Determine the start year and month for the time series
start_year <- year(df_temp$YearMonth[1])
start_month <- month(df_temp$YearMonth[1])

# Create time series objects for log_price and the external regressor
# Frequency is 12 for monthly data
log_price_ts <- ts(df_temp$log_price, start = c(start_year, start_month), frequency = 12)
mean_tavg_ts <- ts(df_temp$MeanTAVG, start = c(start_year, start_month), frequency = 12)

# Determine the total length and the forecast horizon (last 6 months)
n <- length(log_price_ts)
h6 <- 6 # Forecast horizon

# Create training sets using window()
train_log_price <- window(log_price_ts, end = time(log_price_ts)[n - h6])
train_xreg_ts   <- window(mean_tavg_ts, end = time(mean_tavg_ts)[n - h6])

# Create test sets using window()
test_log_price <- window(log_price_ts, start = time(log_price_ts)[n - h6 + 1])
test_xreg_ts   <- window(mean_tavg_ts, start = time(mean_tavg_ts)[n - h6 + 1])
test_dates     <- time(test_log_price) # Get numeric time index for plotting/dataframe

# Convert train_xreg_ts to matrix 
train_xreg_matrix <- as.matrix(train_xreg_ts)
colnames(train_xreg_matrix) <- c("MeanTAVG") 

# Fit the ARIMAX(0,1,1) model 
model_arimax_6m <- tryCatch({
  # Using forecast::Arima
  Arima(train_log_price, order = c(0, 1, 1), xreg = train_xreg_matrix)
}, error = function(e) {
  cat("Error fitting Arima model:", conditionMessage(e), "\n")
  return(NULL)
})

# Forecast on the Test Set ---
if (!is.null(model_arimax_6m)) {

  # Convert test_xreg_ts to matrix 
  test_xreg_matrix <- as.matrix(test_xreg_ts)
  colnames(test_xreg_matrix) <- c("MeanTAVG")
  fc6_arimax <- forecast(model_arimax_6m, h = h6, xreg = test_xreg_matrix)

  # Compute accuracy metrics using forecast::accuracy() on the log scale forecast object
  accuracy6_log <- accuracy(fc6_arimax, test_log_price)
  cat("ARIMAX(1,1,0) 6-Month Forecast Accuracy (Log Scale):\n")
  print(accuracy6_log)

    # Convert forecasts (mean) and actual test values back to the original price scale
  fc6_orig <- exp(fc6_arimax$mean)
  actual6_orig <- exp(test_log_price)

  # Compute error metrics manually on the original scale
  mae_orig <- mean(abs(fc6_orig - actual6_orig))
  rmse_orig <- sqrt(mean((fc6_orig - actual6_orig)^2))
  # Add a small epsilon to actual6_orig denominator to avoid division by zero if necessary
  mape_orig <- mean(abs((fc6_orig - actual6_orig) / (actual6_orig + 1e-10))) * 100

  # Print the results for the original scale
  cat("\nARIMAX(0,1,1) 6-Month Forecast Accuracy on Original Scale:\n")
  cat("MAE:", mae_orig, "\n")
  cat("RMSE:", rmse_orig, "\n")
  cat("MAPE:", mape_orig, "%\n")

# Create a data frame for plogginb
  df6_arimax <- data.frame(
    Month = as.numeric(test_dates), # Use the numeric time index
    Actual = as.numeric(actual6_orig),
    Forecast = as.numeric(fc6_orig)
  )

  # Create the plot 
  p6_arimax <- ggplot(df6_arimax, aes(x = Month)) +
    geom_line(aes(y = Actual, color = "Actual"), linewidth = 1) +
    geom_line(aes(y = Forecast, color = "Forecast"), linewidth = 1, linetype = "dashed") +
    scale_color_manual(name = "Legend", values = c("Actual" = "blue", "Forecast" = "red")) + # Manual colors
    labs(title = "ARIMAX(0,1,1) 6-Month Forecast vs Actual (Original Scale)",
         y = "Price (Original Scale)",
         x = "Time Index") + 
    theme_minimal() +
    theme(legend.position = "top")

  # Print the plot
  print(p6_arimax)

} else {
  cat("Skipping prediction, evaluation, and plotting due to model fitting error.\n")
}
```

## Using ARIMAX (1,1,0) + precipitation for train/test set.
```{r}
# Load necessary libraries so they do not get mixed up 
library(forecast)
library(stats) 
library(ggplot2)
library(lubridate)

# Ensure the data is sorted chronologically by YearMonth
df_prec <- df_prec[order(df_prec$YearMonth), ]

# Determine the start year and month for the time series
start_year <- year(df_prec$YearMonth[1])
start_month <- month(df_prec$YearMonth[1])

# Create time series objects for log_price and the external regressor
# Frequency is 12 for monthly data
log_price_ts <- ts(df_prec$log_price, start = c(start_year, start_month), frequency = 12)
total_prcp_ts <- ts(df_prec$TotalPRCP, start = c(start_year, start_month), frequency = 12)

# Determine the total length and the forecast horizon (last 6 months)
n <- length(log_price_ts)
h6 <- 6 # Forecast horizon

# Create training sets using window()
train_log_price <- window(log_price_ts, end = time(log_price_ts)[n - h6])
train_xreg_ts   <- window(total_prcp_ts, end = time(total_prcp_ts)[n - h6])

# Create test sets using window()
test_log_price <- window(log_price_ts, start = time(log_price_ts)[n - h6 + 1])
test_xreg_ts   <- window(total_prcp_ts, start = time(total_prcp_ts)[n - h6 + 1])
test_dates     <- time(test_log_price) # Get numeric time index for plotting/dataframe

# Convert train_xreg_ts to matrix 
train_xreg_matrix <- as.matrix(train_xreg_ts)
colnames(train_xreg_matrix) <- c("TotalPRCP") 

# Fit the ARIMAX(1,1,0) model 
model_arimax_6m <- tryCatch({
  # Using forecast::Arima
  Arima(train_log_price, order = c(1, 1, 0), xreg = train_xreg_matrix)
}, error = function(e) {
  cat("Error fitting Arima model:", conditionMessage(e), "\n")
  return(NULL)
})

# Forecast on the Test Set ---
if (!is.null(model_arimax_6m)) {

  # Convert test_xreg_ts to matrix 
  test_xreg_matrix <- as.matrix(test_xreg_ts)
  colnames(test_xreg_matrix) <- c("TotalPRCP")
  fc6_arimax <- forecast(model_arimax_6m, h = h6, xreg = test_xreg_matrix)

  # Compute accuracy metrics using forecast::accuracy() on the log scale forecast object
  accuracy6_log <- accuracy(fc6_arimax, test_log_price)
  cat("ARIMAX(1,1,0) 6-Month Forecast Accuracy (Log Scale):\n")
  print(accuracy6_log)

    # Convert forecasts (mean) and actual test values back to the original price scale
  fc6_orig <- exp(fc6_arimax$mean)
  actual6_orig <- exp(test_log_price)

  # Compute error metrics manually on the original scale
  mae_orig <- mean(abs(fc6_orig - actual6_orig))
  rmse_orig <- sqrt(mean((fc6_orig - actual6_orig)^2))
  # Add a small epsilon to actual6_orig denominator to avoid division by zero if necessary
  mape_orig <- mean(abs((fc6_orig - actual6_orig) / (actual6_orig + 1e-10))) * 100

  # Print the results for the original scale
  cat("\nARIMAX(1,1,0) 6-Month Forecast Accuracy on Original Scale:\n")
  cat("MAE:", mae_orig, "\n")
  cat("RMSE:", rmse_orig, "\n")
  cat("MAPE:", mape_orig, "%\n")

# Create a data frame for plogginb
  df6_arimax <- data.frame(
    Month = as.numeric(test_dates), # Use the numeric time index
    Actual = as.numeric(actual6_orig),
    Forecast = as.numeric(fc6_orig)
  )

  # Create the plot 
  p6_arimax <- ggplot(df6_arimax, aes(x = Month)) +
    geom_line(aes(y = Actual, color = "Actual"), linewidth = 1) +
    geom_line(aes(y = Forecast, color = "Forecast"), linewidth = 1, linetype = "dashed") +
    scale_color_manual(name = "Legend", values = c("Actual" = "blue", "Forecast" = "red")) + # Manual colors
    labs(title = "ARIMAX(1,1,0) 6-Month Forecast vs Actual (Original Scale)",
         y = "Price (Original Scale)",
         x = "Time Index") + 
    theme_minimal() +
    theme(legend.position = "top")

  # Print the plot
  print(p6_arimax)

} else {
  cat("Skipping prediction, evaluation, and plotting due to model fitting error.\n")
}
```

## Using ARIMAX (0,1,1) + precipitation for train/test set.
```{r}
# Load necessary libraries so they do not get mixed up 
library(forecast)
library(stats) 
library(ggplot2)
library(lubridate)

# Ensure the data is sorted chronologically by YearMonth
df_prec <- df_prec[order(df_prec$YearMonth), ]

# Determine the start year and month for the time series
start_year <- year(df_prec$YearMonth[1])
start_month <- month(df_prec$YearMonth[1])

# Create time series objects for log_price and the external regressor
# Frequency is 12 for monthly data
log_price_ts <- ts(df_prec$log_price, start = c(start_year, start_month), frequency = 12)
total_prcp_ts <- ts(df_prec$TotalPRCP, start = c(start_year, start_month), frequency = 12)

# Determine the total length and the forecast horizon (last 6 months)
n <- length(log_price_ts)
h6 <- 6 # Forecast horizon

# Create training sets using window()
train_log_price <- window(log_price_ts, end = time(log_price_ts)[n - h6])
train_xreg_ts   <- window(total_prcp_ts, end = time(total_prcp_ts)[n - h6])

# Create test sets using window()
test_log_price <- window(log_price_ts, start = time(log_price_ts)[n - h6 + 1])
test_xreg_ts   <- window(total_prcp_ts, start = time(total_prcp_ts)[n - h6 + 1])
test_dates     <- time(test_log_price) # Get numeric time index for plotting/dataframe

# Convert train_xreg_ts to matrix 
train_xreg_matrix <- as.matrix(train_xreg_ts)
colnames(train_xreg_matrix) <- c("TotalPRCP") 

# Fit the ARIMAX(0,1,1) model 
model_arimax_6m <- tryCatch({
  # Using forecast::Arima
  Arima(train_log_price, order = c(0, 1, 1), xreg = train_xreg_matrix)
}, error = function(e) {
  cat("Error fitting Arima model:", conditionMessage(e), "\n")
  return(NULL)
})

# Forecast on the Test Set ---
if (!is.null(model_arimax_6m)) {

  # Convert test_xreg_ts to matrix 
  test_xreg_matrix <- as.matrix(test_xreg_ts)
  colnames(test_xreg_matrix) <- c("TotalPRCP")
  fc6_arimax <- forecast(model_arimax_6m, h = h6, xreg = test_xreg_matrix)

  # Compute accuracy metrics using forecast::accuracy() on the log scale forecast object
  accuracy6_log <- accuracy(fc6_arimax, test_log_price)
  cat("ARIMAX(1,1,0) 6-Month Forecast Accuracy (Log Scale):\n")
  print(accuracy6_log)

    # Convert forecasts (mean) and actual test values back to the original price scale
  fc6_orig <- exp(fc6_arimax$mean)
  actual6_orig <- exp(test_log_price)

  # Compute error metrics manually on the original scale
  mae_orig <- mean(abs(fc6_orig - actual6_orig))
  rmse_orig <- sqrt(mean((fc6_orig - actual6_orig)^2))
  # Add a small epsilon to actual6_orig denominator to avoid division by zero if necessary
  mape_orig <- mean(abs((fc6_orig - actual6_orig) / (actual6_orig + 1e-10))) * 100

  # Print the results for the original scale
  cat("\nARIMAX(0,1,1) 6-Month Forecast Accuracy on Original Scale:\n")
  cat("MAE:", mae_orig, "\n")
  cat("RMSE:", rmse_orig, "\n")
  cat("MAPE:", mape_orig, "%\n")

# Create a data frame for plogginb
  df6_arimax <- data.frame(
    Month = as.numeric(test_dates), # Use the numeric time index
    Actual = as.numeric(actual6_orig),
    Forecast = as.numeric(fc6_orig)
  )

  # Create the plot 
  p6_arimax <- ggplot(df6_arimax, aes(x = Month)) +
    geom_line(aes(y = Actual, color = "Actual"), linewidth = 1) +
    geom_line(aes(y = Forecast, color = "Forecast"), linewidth = 1, linetype = "dashed") +
    scale_color_manual(name = "Legend", values = c("Actual" = "blue", "Forecast" = "red")) + # Manual colors
    labs(title = "ARIMAX(0,1,1) 6-Month Forecast vs Actual (Original Scale)",
         y = "Price (Original Scale)",
         x = "Time Index") + 
    theme_minimal() +
    theme(legend.position = "top")

  # Print the plot
  print(p6_arimax)

} else {
  cat("Skipping prediction, evaluation, and plotting due to model fitting error.\n")
}
```

## Using ARIMAX (6,1,7) + precipitation for train/test set.
```{r}
# Load necessary libraries so they do not get mixed up 
library(forecast)
library(stats) 
library(ggplot2)
library(lubridate)

# Ensure the data is sorted chronologically by YearMonth
df_prec <- df_prec[order(df_prec$YearMonth), ]

# Determine the start year and month for the time series
start_year <- year(df_prec$YearMonth[1])
start_month <- month(df_prec$YearMonth[1])

# Create time series objects for log_price and the external regressor
# Frequency is 12 for monthly data
log_price_ts <- ts(df_prec$log_price, start = c(start_year, start_month), frequency = 12)
total_prcp_ts <- ts(df_prec$TotalPRCP, start = c(start_year, start_month), frequency = 12)

# Determine the total length and the forecast horizon (last 6 months)
n <- length(log_price_ts)
h6 <- 6 # Forecast horizon

# Create training sets using window()
train_log_price <- window(log_price_ts, end = time(log_price_ts)[n - h6])
train_xreg_ts   <- window(total_prcp_ts, end = time(total_prcp_ts)[n - h6])

# Create test sets using window()
test_log_price <- window(log_price_ts, start = time(log_price_ts)[n - h6 + 1])
test_xreg_ts   <- window(total_prcp_ts, start = time(total_prcp_ts)[n - h6 + 1])
test_dates     <- time(test_log_price) # Get numeric time index for plotting/dataframe

# Convert train_xreg_ts to matrix 
train_xreg_matrix <- as.matrix(train_xreg_ts)
colnames(train_xreg_matrix) <- c("TotalPRCP") 

# Fit the ARIMAX(6,1,7) model 
model_arimax_6m <- tryCatch({
  # Using forecast::Arima
  Arima(train_log_price, order = c(6, 1, 7), xreg = train_xreg_matrix)
}, error = function(e) {
  cat("Error fitting Arima model:", conditionMessage(e), "\n")
  return(NULL)
})

# Forecast on the Test Set ---
if (!is.null(model_arimax_6m)) {

  # Convert test_xreg_ts to matrix 
  test_xreg_matrix <- as.matrix(test_xreg_ts)
  colnames(test_xreg_matrix) <- c("TotalPRCP")
  fc6_arimax <- forecast(model_arimax_6m, h = h6, xreg = test_xreg_matrix)

  # Compute accuracy metrics using forecast::accuracy() on the log scale forecast object
  accuracy6_log <- accuracy(fc6_arimax, test_log_price)
  cat("ARIMAX(6,1,7) 6-Month Forecast Accuracy (Log Scale):\n")
  print(accuracy6_log)

    # Convert forecasts (mean) and actual test values back to the original price scale
  fc6_orig <- exp(fc6_arimax$mean)
  actual6_orig <- exp(test_log_price)

  # Compute error metrics manually on the original scale
  mae_orig <- mean(abs(fc6_orig - actual6_orig))
  rmse_orig <- sqrt(mean((fc6_orig - actual6_orig)^2))
  # Add a small epsilon to actual6_orig denominator to avoid division by zero if necessary
  mape_orig <- mean(abs((fc6_orig - actual6_orig) / (actual6_orig + 1e-10))) * 100

  # Print the results for the original scale
  cat("\nARIMAX(6,1,7) 6-Month Forecast Accuracy on Original Scale:\n")
  cat("MAE:", mae_orig, "\n")
  cat("RMSE:", rmse_orig, "\n")
  cat("MAPE:", mape_orig, "%\n")

# Create a data frame for plogginb
  df6_arimax <- data.frame(
    Month = as.numeric(test_dates), # Use the numeric time index
    Actual = as.numeric(actual6_orig),
    Forecast = as.numeric(fc6_orig)
  )

  # Create the plot 
  p6_arimax <- ggplot(df6_arimax, aes(x = Month)) +
    geom_line(aes(y = Actual, color = "Actual"), linewidth = 1) +
    geom_line(aes(y = Forecast, color = "Forecast"), linewidth = 1, linetype = "dashed") +
    scale_color_manual(name = "Legend", values = c("Actual" = "blue", "Forecast" = "red")) + # Manual colors
    labs(title = "ARIMAX(6,1,7) with Precipitation 6-Month Forecast vs Actual",
         y = "Price (Original Scale)",
         x = "Time Index") + 
    theme_minimal() +
    theme(legend.position = "top")

  # Print the plot
  print(p6_arimax)

} else {
  cat("Skipping prediction, evaluation, and plotting due to model fitting error.\n")
}
```

